---
title: "PLCO Lung Analysis"
author: "James Long"
always_allow_html: yes
date: "`r Sys.Date()`"
bibliography: refs.bib
output:
  html_document:
    theme: united
    toc: yes
    toc_collapsed: yes
    toc_float: yes
---

```{r alternative-preambles,echo=FALSE,include=FALSE,eval=FALSE}
output:
  html_document:
    theme: united
    toc: yes
    toc_collapsed: yes
    toc_float: yes



## if using word replace output with
output:
  word_document:
    toc: yes

## if using pdf replace output with
output:
  pdf_document:
    toc: yes
header-includes:
  - \usepackage{xcolor}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

## colorizes text when using html or pdf output
## just spits out text x for word
colorize <- function(x, color) {
  if(color=="todo"){
    color <- "red"
    x <- paste0("TODO: ",x)
  }
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}

todo <- function(x){
  paste0("**",colorize(x,"todo"),"**")
}

set.seed(20230126)
library(gtsummary)
library(ggplot2)
library(survival)
library(survminer)
library(kableExtra)
library(randomForestSRC)
library(cowplot)
#library(huxtable)

outf <- "20230126-plco-lung-report/"
unlink(outf,recursive=TRUE)
dir.create(outf,recursive=TRUE)
```


```{r}
## load data and recode several variables
fname <- "../data/package-plco-808.2022-12-25/Lung/lung_data_mar22_d032222.csv"
dat <- read.csv(fname)


#table(dat$arm,useNA="always")

dat$armf <- NA_character_
dat$armf[dat$arm==1] <- "Intervention"
dat$armf[dat$arm==2] <- "Control"

dat$center <- as.factor(dat$center)
dat$fstcan_exitstat <- as.factor(dat$fstcan_exitstat)


## category names for type of detection
temp <- rep(NA_character_,nrow(dat))
temp[dat$intstatl_cat==0] <- "No Cancer"
temp[dat$intstatl_cat==1] <- "Control"
temp[dat$intstatl_cat==2] <- "Other"
temp[dat$intstatl_cat==3] <- "Other"
temp[dat$intstatl_cat==4] <- "Interval"
temp[dat$intstatl_cat==5] <- "Screen"
dat$Detection <- factor(temp,
                       levels=c("Screen","Interval",
                                "Control","Other",
                                "No Cancer"))
#table(dat$Detection,dat$intstatl_cat)

## category names for age level
temp <- rep(NA_character_,nrow(dat))
temp[dat$agelevel==0] <- "<=59"
temp[dat$agelevel==1] <- "60-64"
temp[dat$agelevel==2] <- "65-69"
temp[dat$agelevel==3] <- ">=70"
temp <- factor(temp,levels=c("<=59","60-64","65-69",">=70"))
dat$Age <- temp


## recode lung hist type category
recode <- c("Small cell carcinoma",
            "Adenocarcinoma",
            "Bronchiolo-alveolar",
            "Squamous cell",
            "Large cell",
            "Other NSC",
            "Carcinoma, NOS",
            "Other/Unknown")
names(recode) <- as.character(c(1,2,3,4,5,6,8,9))
temp <- recode[as.character(dat$lung_histtype_cat)]
temp <- factor(temp,levels=recode)
#table(temp,dat$lung_histtype_cat,useNA="always")
#table(temp,dat$lung_cancer,useNA="always")
dat$Histology <- temp

## category names for lung cancer (yes/no)
temp <- rep(NA_character_,nrow(dat))
temp[dat$lung_cancer==1] <- "Yes"
temp[dat$lung_cancer==0] <- "No"
dat$lung_cancer <- as.factor(temp)



#table(dat$Detection,dat$intstatl_cat)
#table(dat$Detection,dat$arm)
#table(dat$Detection,dat$lung_cancer)


## make stage a factor
temp <- rep(NA_character_,nrow(dat))
temp[dat$lung_stage<150] <- "Stage I"
temp[dat$lung_stage>150 & dat$lung_stage < 250] <- "Stage II"
temp[dat$lung_stage>250 & dat$lung_stage < 350] <- "Stage III"
temp[dat$lung_stage==400] <- "Stage IV"
temp[dat$lung_stage==993] <- "Small Cell"
temp[is.na(dat$lung_stage) & dat$lung_cancer=="Yes"] <- "Unknown"
#table(temp,dat$lung_stage,useNA="always")
dat$Stage <- factor(temp,levels=c("Stage I","Stage II","Stage III","Stage IV","Small Cell","Unknown"))
#table(dat$Stage,dat$lung_stage,useNA="always")




## update sex variable
temp <- rep(NA_character_,nrow(dat))
temp[dat$sex==1] <- "Male"
temp[dat$sex==2] <- "Female"
#table(dat$sex,temp)
dat$sex <- factor(temp)
colnames(dat)[colnames(dat)=="sex"] <- "Sex"


recode <- c("Grade I",
            "Grade II",
            "Grade III",
            "Grade IV",
            "Other/Unknown",
            "Other/Unknown")
names(recode) <- as.character(c(1,2,3,4,5,9))
temp <- recode[dat$lung_grade]
temp[dat$lung_cancer=="Yes" & is.na(temp)] <- "Other/Unknown"
dat$Grade <- temp
#table(dat$Grade,dat$lung_grade,useNA="always")
#table(dat$Grade,dat$lung_cancer,useNA="always")




# get cigarette smoking
#table(dat$cig_stat,useNA="always")
Smoked <- rep(NA_character_,nrow(dat))
Smoked[dat$cig_stat==0] <- "No"
Smoked[dat$cig_stat==1] <- "Yes"
Smoked[dat$cig_stat==2] <- "Yes"
Smoked[is.na(dat$cig_stat)] <- "Unknown"
dat$Smoked <- factor(Smoked,levels=c("No","Yes","Unknown"))
#table(dat$cig_stat,dat$Smoked,useNA="always")

# compute number of days from detection to death
# will be missing for patients who did not have
# lung cancer (very common)
dat$DD_years <- (dat$mortality_exitdays-dat$lung_cancer_diagdays)/365.25


## everyone is known to be alive or dead
## if alive, reason for death is always missing
## id dead, rarely is cause of death missing
#table(dat$d_cancersite,dat$is_dead,useNA="always")
#table(dat$d_cancersite==2,dat$lung_cancer,useNA="always")
#table(dat$d_cancersite[dat$lung_cancer=="Yes"],useNA="always")


## create variable which indicates if death is due to lung cancer
died_lung_cancer <- rep("Other Death",nrow(dat))
died_lung_cancer[dat$is_dead==0] <- "Alive"
died_lung_cancer[dat$d_cancersite==2] <- "Lung Cancer Death"
dat$died_lung_cancer <- died_lung_cancer


## x-axis limits for survival plots
xlims <- c(0,10)
break.x.by <- 2


## define custom test so that can compute p-values for categorical variables
## when one category level has limited data (so asymptotic not appropriate)
## and another category level has a lot of data (so exact computations too expensive)
## see: https://stackoverflow.com/questions/61360954/error-in-add-p-for-variable-x-and-test-fisher-test-p-value-omitted
fisher.test.simulate.p.values <- function(data, variable, by, ...) {
  result <- list()
  test_results <- stats::fisher.test(data[[variable]], data[[by]], simulate.p.value = TRUE)
  result$p <- test_results$p.value
  result$test <- test_results$method
  result
}
```

```{r}
### determine longest follow up time
summary(dat$DD_years)


summary(dat$lung_cancer_diagdays/365.25)
```


## Intervention versus Control Demographic Table

```{r message=FALSE}
add_p_ex1 <-
  dat[c("Age","armf","Sex","lung_cancer","Detection","Stage","Grade","pack_years","Smoked")] %>%
  tbl_summary(by = armf) %>%
  add_p(test = list(all_categorical() ~ "fisher.test.simulate.p.values"))    
add_p_ex1
```

### Follow up

```{r,fig.height=4,fig.width=4,fig.align='center'}
hist(dat$mortality_exitdays,xlab="Follow up Time (Days)",main="")
```

The median follow up time is `r round(median(dat$mortality_exitdays))` days.

### Overall Survival

Similar overall survival between the two groups.

```{r}
fit  <- survfit(Surv(mortality_exitdays,is_dead)~armf,data=dat)
p <- ggsurvplot(fit,data=dat,risk.table=TRUE,ylim=c(0,1),legend.labs=levels(dat$armf),
                legend.title="",pval=TRUE,legend="none",xlim=c(0,4000),break.x.by=1000)
p[[1]] <- p[[1]] + theme_grey() + theme(legend.position="none") + xlab("Time (days)")
print(p)
```



```{r}
##### cohort: all patients diagnosed with NSCLC with staging
## select patients with lung cancer
dat_sub <- dat[dat$lung_cancer=="Yes",]
## select patients with known stage
dat_sub <- dat_sub[dat_sub$Stage!="Unknown",]
## select patients who do not have small cell
dat_sub <- dat_sub[!(dat_sub$lung_cancer_type==2),]



dat_sub$Stage <- factor(as.character(dat_sub$Stage),
                        levels=c("Stage I","Stage II","Stage III","Stage IV"))
dat_sub$Detection <- factor(as.character(dat_sub$Detection),
                     levels=c("Screen","Interval","Other","Control"))

levs <- levels(dat_sub$Histology)
levs <- levs[levs %in% dat_sub$Histology]
dat_sub$Histology <- factor(as.character(dat_sub$Histology),levels=levs)


## features for modeling
resp <- "Surv(DD_years,is_dead)"
feats <- c("Detection","Stage","Age","Sex","Smoked","Histology")
```


## Cancer Cases

We now subset to patients with Non-small cell lung cancer (NSCLC) who have Stage information. The variable Detection takes a value based on how the cancer was detected.

* Screen: Patient was in intervention arm and tumor was detected during screening
* Interval: Patient was in intervention arm and tumor was detected between screenings
* Other: Other Intervention includes all patients diagnosed in Intervention Arm who are not in the Screen or Interval Category. This includes patients who
    * Never attended any screenings
    * Were diagnosed with cancer after screening period ended (Post)
* Control: Patient was in control arm

There is evidence of phase shift when one compares the distribution of Stage I vs. IV cancers in Screen detected versus other detection methods.

### Lung Cancer Deaths

Among patients diagnosed with lung cancer, we compute the fraction that died from lung cancer, fraction died from other causes, and the fraction still alive.

```{r}
tab <- table(dat_sub$died_lung_cancer)/nrow(dat_sub)
tab <- as.data.frame(tab)
colnames(tab) <- c("Cause of Death","Fraction")
kbl(tab) %>%
  kable_paper(bootstrap_options="striped",full_width = F)
```


### Lung Cancer Case Summary

```{r message=FALSE}
add_p_ex1 <-
  dat_sub[feats] %>%
  tbl_summary(by = Detection) %>%
  add_p(test = list(all_categorical() ~ "fisher.test.simulate.p.values"))
add_p_ex1

## create huxtable for output
ht <- as_hux_table(add_p_ex1)
huxtable::bottom_padding(ht) <- 0
huxtable::top_padding(ht) <- 0
huxtable::right_padding(ht) <- 0
#huxtable::left_padding(ht) <- 0
huxtable::caption(ht) <- "Characteristics of Lung cancers detected in PLCO by Detection Method for Extended Cohort."
huxtable::label(ht) <- "tab:table1"
huxtable::number_format(ht) <- NA

## convert to latex
ht_latex <- ht %>% huxtable::to_latex()
## remove some formatting
ht_latex <- gsub("```{=latex}","",ht_latex,fixed=TRUE)
ht_latex <- gsub("```","",ht_latex,fixed=TRUE)
cat(ht_latex,file=paste0(outf,"table1.tex"))
```



### Lung Cancer Survival

Survival is better for screen detected.


```{r fig.height=6.5,fig.width=7,align="center"}
fit  <- survfit(Surv(DD_years,is_dead)~Detection,data=dat_sub)
p1 <- ggsurvplot(fit,data=dat_sub,risk.table=TRUE,ylim=c(0,1),legend.labs=levels(dat_sub$Detection),
                legend.title="",pval = TRUE,legend="none",pval.coord=c(6,0.8),
                xlim=xlims,break.x.by=break.x.by)
p1[[1]] <- p1[[1]] + theme_grey() + theme(legend.position="top") + xlab("Time (years)")
print(p1)

pdf(paste0(outf,"km-overall.pdf"),width=6,height=7,onefile=FALSE)
print(p1)
invisible(dev.off())
```


### Lung Cancer Survival by Stage

Survival is better for screen detected after controlling for stage.

```{r fig.width=6,fig.height=6,message=FALSE,warning=FALSE,fig.align="center"}
fit  <- survfit(Surv(DD_years,is_dead)~Detection,data=dat_sub)
p2 <- ggsurvplot_facet(fit,data=dat_sub,risk.table=TRUE,ylim=c(0,1),short.panel.labs=TRUE,
                legend.title="",pval = TRUE,facet.by="Stage",pval.coord=c(6,0.8),
                break.x.by=break.x.by,xlim=xlims)
p2 <- p2 + theme_grey() + theme(legend.position="top") + xlab("Time (years)")
print(p2)

pdf(paste0(outf,"km-bystage.pdf"),width=6,height=6)
print(p2)
invisible(dev.off())
```

```{r}
p <- plot_grid(p1$plot, p2, labels = c('a)', 'b)'), label_size = 12)
pdf(paste0(outf,"km-merged.pdf"),width=10,height=5)
print(p)
invisible(dev.off())
```

```{r}
## what fraction of lung cancer patients die of disease
table(dat_sub$died_lung_cancer)/nrow(dat_sub)
```

### Prediction Models

Fit Cox PH model to predict survival. Detection Method is significant even after controlling for age, stage, histology, etc.

```{r}
fm <- as.formula(paste0(resp,"~",paste0(feats,collapse="+")))
fit <- coxph(fm,data=dat_sub)
out <- tbl_regression(fit)
out


ht <- as_hux_table(out)
huxtable::bottom_padding(ht) <- 0
huxtable::top_padding(ht) <- 0
huxtable::right_padding(ht) <- 0
huxtable::top_padding(ht) <- 0
huxtable::caption(ht) <- "Multivariate Cox PH model to predict overall survival following diagnosis with lung cancer. Model is fit on Extended Cohort."
huxtable::label(ht) <- "tab:coxph-fit"


## convert to latex
ht_latex <- ht %>% huxtable::to_latex()
## remove some formatting
ht_latex <- gsub("```{=latex}","",ht_latex,fixed=TRUE)
ht_latex <- gsub("```","",ht_latex,fixed=TRUE)
cat(ht_latex,file=paste0(outf,"coxph-fit.tex"))
```

Assess Concordance Index of Model

```{r}
concordance(fit)
```


### Quantify Strength of Detection Method as Feature

To assess the importance of detection method relative to other features in predicting survival, the following procedure was performed:

* Compute Concordance Index $\widehat{C}$ for full model above
* For each variable $X$ in the model:
    * Fit submodel which uses all variables except for $X$ and determine concordance index $\widehat{C}_{-X}$
    * Compute $\widehat{\Delta}_X = \widehat{C}-\widehat{C}_{-X}$, the change in concordance when feature $X$ is added to the model
    * Compute p-value for difference in concordances
        * $H_0: \Delta_{X} \equiv C - C_{-X} = 0$

```{r}
bmat <- matrix(c(1,-1),ncol=1)
res <- matrix(NA_real_,nrow=length(feats),ncol=2)
for(ii in 1:length(feats)){
  fm_sub <- as.formula(paste0(resp,"~",paste0(feats[-ii],collapse="+")))
  fit_sub <- coxph(fm_sub,data=dat_sub)
  a <- concordance(fit,fit_sub)
  se <- sqrt(t(bmat)%*%a$var%*%bmat)
  zstat <- diff(a$concordance)/se
  res[ii,1] <- diff(rev(a$concordance))
  res[ii,2] <- 2*pnorm(-abs(zstat))
}

res <- data.frame(Features=feats,DeltaCIndex=res[,1],pvalue=res[,2])


res <- res[order(res[,3]),]
res[,3] <- scales::pvalue(res[,3],accuracy=0.00001)
kbl(res,row.names=FALSE) %>%
  kable_paper(bootstrap_options="basic",full_width = F)
```




```{r}
## fit survival forests to data
## results reported later in output
fm <- as.formula(paste0(resp,"~",paste0(feats,collapse="+")))
fit <- rfsrc(fm,data=dat_sub,importance="permute")
imp_full <- fit$importance
```







## Cancer Cases within 4 Years

The previous analysis is repeated but restricting analysis to lung cancers diagnosed within 4 years of randomization. These 4 years include 3 years of screening (4 total screenings at years 0,1,2 and 3) in the intervention group followed by 1 year of follow up.


```{r}
dat_sub <- dat_sub[dat_sub$lung_cancer_diagdays<365.25*4,]
```

### Lung Cancer Deaths

Among patients diagnosed with lung cancer, we compute the fraction that died from lung cancer, fraction died from other causes, and the fraction still alive.

```{r}
tab <- table(dat_sub$died_lung_cancer)/nrow(dat_sub)
tab <- as.data.frame(tab)
colnames(tab) <- c("Cause of Death","Fraction")
kbl(tab) %>%
  kable_paper(bootstrap_options="striped",full_width = F)
```




### Lung Cancer Case Summary

```{r message=FALSE}
add_p_ex1 <-
  dat_sub[feats] %>%
  tbl_summary(by = Detection) %>%
  add_p(test = list(all_categorical() ~ "fisher.test.simulate.p.values"))  
add_p_ex1


## create huxtable for output
ht <- as_hux_table(add_p_ex1)
huxtable::bottom_padding(ht) <- 0
huxtable::top_padding(ht) <- 0
huxtable::right_padding(ht) <- 0
huxtable::caption(ht) <- "Characteristics of Lung cancers detected in PLCO by Detection Method."
huxtable::label(ht) <- "tab:table1-within4"

## convert to latex
ht_latex <- ht %>% huxtable::to_latex()
## remove some formatting
ht_latex <- gsub("```{=latex}","",ht_latex,fixed=TRUE)
ht_latex <- gsub("```","",ht_latex,fixed=TRUE)
cat(ht_latex,file=paste0(outf,"table1-within4.tex"))
```



### Lung Cancer Survival


```{r fig.height=6.5,fig.width=7,align="center"}
fit  <- survfit(Surv(DD_years,is_dead)~Detection,data=dat_sub)
p1 <- ggsurvplot(fit,data=dat_sub,risk.table=TRUE,ylim=c(0,1),legend.labs=levels(dat_sub$Detection),
                legend.title="",pval = TRUE,legend="none",pval.coord=c(6,0.8),
                xlim=xlims,break.x.by=break.x.by)
p1[[1]] <- p1[[1]] + theme_grey() + theme(legend.position="top") + xlab("Time (years)")
print(p1)

pdf(paste0(outf,"km-overall-within4.pdf"),width=6,height=7,onefile=FALSE)
print(p1)
invisible(dev.off())
```


### Lung Cancer Survival by Stage

Survival is much better for screen detected after controlling for stage.

```{r fig.width=6,fig.height=6,message=FALSE,warning=FALSE,fig.align="center"}
fit  <- survfit(Surv(DD_years,is_dead)~Detection,data=dat_sub)
p2 <- ggsurvplot_facet(fit,data=dat_sub,risk.table=TRUE,ylim=c(0,1),short.panel.labs=TRUE,
                legend.title="",pval = TRUE,facet.by="Stage",pval.coord=c(6,0.8),
                break.x.by=break.x.by,xlim=xlims)
p2 <- p2 + theme_grey() + theme(legend.position="top") + xlab("Time (years)")
print(p2)

pdf(paste0(outf,"km-bystage-within4.pdf"),width=6,height=6)
print(p2)
invisible(dev.off())
```

```{r}
## determine median by detection method for stage 1
fit  <- survfit(Surv(DD_years,is_dead)~Detection,data=dat_sub,subset=Stage=="Stage I")
fit
```

```{r}
p <- plot_grid(p1$plot, p2, labels = c('a)', 'b)'), label_size = 12)
pdf(paste0(outf,"km-merged-within4.pdf"),width=10,height=5)
print(p)
invisible(dev.off())
```


### Prediction Models

Fit Cox PH model to predict survival. Detection Method is significant even after controlling for age, stage, histology, etc.

```{r}
## delete observations with any missingness in these variables
fm <- as.formula(paste0(resp,"~",paste0(feats,collapse="+")))
fit <- coxph(fm,data=dat_sub)
out <- tbl_regression(fit)
out


ht <- as_hux_table(out)
huxtable::bottom_padding(ht) <- 0
huxtable::top_padding(ht) <- 0
huxtable::right_padding(ht) <- 0
huxtable::top_padding(ht) <- 0
huxtable::caption(ht) <- "Multivariate Cox PH model to predict overall survival following diagnosis with lung cancer."
huxtable::label(ht) <- "tab:coxph-fit-within4"


## convert to latex
ht_latex <- ht %>% huxtable::to_latex()
## remove some formatting
ht_latex <- gsub("```{=latex}","",ht_latex,fixed=TRUE)
ht_latex <- gsub("```","",ht_latex,fixed=TRUE)
cat(ht_latex,file=paste0(outf,"coxph-fit-within4.tex"))
```

Assess Concordance Index of Model

```{r}
concordance(fit)
```







### Quantify Strength of Detection Method as Feature

To assess the importance of detection method relative to other features in predicting survival, the following procedure was performed:

* Compute Concordance Index $\widehat{C}$ for full model above
* For each variable $X$ in the model:
    * Fit submodel which uses all variables except for $X$ and determine concordance index $\widehat{C}_{-X}$
    * Compute $\widehat{\Delta}_X = \widehat{C}-\widehat{C}_{-X}$, the change in concordance when feature $X$ is added to the model
    * Compute p-value for difference in concordances
        * $H_0: \Delta_{X} \equiv C - C_{-X} = 0$

```{r}
bmat <- matrix(c(1,-1),ncol=1)
res <- matrix(NA_real_,nrow=length(feats),ncol=2)
for(ii in 1:length(feats)){
  fm_sub <- as.formula(paste0(resp,"~",paste0(feats[-ii],collapse="+")))
  fit_sub <- coxph(fm_sub,data=dat_sub)
  a <- concordance(fit,fit_sub)
  se <- sqrt(t(bmat)%*%a$var%*%bmat)
  zstat <- diff(a$concordance)/se
  res[ii,1] <- diff(rev(a$concordance))
  res[ii,2] <- 2*pnorm(-abs(zstat))
}

res <- data.frame(Features=feats,DeltaCIndex=res[,1],pvalue=res[,2])


res <- res[order(res[,3]),]
res[,3] <- scales::pvalue(res[,3],accuracy=0.00001)
kbl(res,row.names=FALSE) %>%
  kable_paper(bootstrap_options="basic",full_width = F)
```




## Survival Forests

Random Survival Forests are fit to the full and restricted cohorts. Variable importance measures are computed to assess how important method of detection is relative to other prognostic variables.

```{r}
fm <- as.formula(paste0(resp,"~",paste0(feats,collapse="+")))
fit <- rfsrc(fm,data=dat_sub,importance="permute")
imp_sub <- fit$importance
```



```{r}
res <- cbind("4-Year Cohort"=imp_sub,"Extended Cohort"=imp_full)
res <- round(res,5)
res <- res[order(res[,1],decreasing=TRUE),]
kbl(res) %>%
  kable_paper(bootstrap_options="striped",full_width = F)
```

```{r}
## make nice output
caption <- "Permutation Variable Importance in Random Survival Forests Models. Detection is the second most important variable, after Stage, in the 4-Year Cohort. In the Extended Cohort, Detection is more important than both Sex and Smoked. These results show that method of detection has independent prognostic value in models other than CoxPH.\\label{tab:varimp}"
tab <- kbl(res,booktabs=TRUE,format="latex",escape=FALSE,caption=caption)
save_kable(tab,file=paste0(outf,"varimp.tex"))
```


